---
title: "GO Analysis"
author: "Brennan McDonald"
date: "2024-04-03"
output: html_document
---

# Overview

This script provides the code needed to perform the Gene Ontology (GO) over representation analysis for the genes expressed in the different Mfuzz clusters (which requires running "mfuzz-pipeline.Rmd" first). The plots generated in this script appear in Figure 7.

```{r}
# Load packages
library(clusterProfiler)
library(org.Herythrogramma.eg.db)
library(tidyverse)
library(enrichplot)
```


# GO analysis comparing genes in Mfuzz expression clusters

```{r}
# Load the expression profile assignments for each gene from the Mfuzz pipeline
# This table was generated in the "mfuzz-pipeline.Rmd" script
cluster_inputs <- read.csv("6-60hr Integrated UMAP/pseudobulk/gene_cluster_assignments.csv")
#cluster_inputs <- read.csv("6-60hr Integrated UMAP/pseudobulk/go_cluster_inputs.csv")

cluster_inputs %>% 
  group_by(clusters) %>% 
  summarize(n = n())
```


This code uses the GO over representation analysis pipeline provided by the `clusterProfiler` package. The plots are generated using functions from the `enrichplot` package. The genes expressed in each of the Mfuzz clusters (9 total) are compared to the genes expressed in the other clusters as a background set.

```{r}
# This generates a list, containing vectors with the genes found in each Mfuzz cluster
cluster_list <- list()
for (c in 1:9) {
  genes <- cluster_inputs %>% 
    filter(clusters == c) %>% 
    dplyr::select(GID)
  name <- paste("c", as.character(c), sep = "")
  cluster_list[[name]] <- genes[,1]
}

# This perform the GO over representation analysis by comparing the genes in each
# Mfuzz cluster against one another, looking for enriched Biological Processes
cl_compare <- compareCluster(geneClusters = cluster_list,
                             fun = "enrichGO",
                             OrgDb = org.Herythrogramma.eg.db,
                             keyType = "GID",
                             ont = "BP", # Looks for enriched Biological Processes
                             pvalueCutoff = 0.05,
                             pAdjustMethod = "BH",
                             qvalueCutoff = 0.2)


# This generates the heatmap shown in Figure 7.
simp_cl_compare <- gofilter(cl_compare, level = 4)
simp_cl_similarities <- pairwise_termsim(simp_cl_compare)
treeplot(simp_cl_similarities, showCategory = 4)
```





